Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-11-01 response_to_reviewers.Rmd

> For Section 6, there is also now a "Big Data Derby" in horse racing.

We thank the reviewer for this suggestion. We have added a reference to it.

> All in all, I think the article explains the ideas well and covers a lot of ground, but more care should be taken towards the organization and what to include as figures. The last half of the paper has a very different pace than the first half. I think a common theme in the paper is the rise in tracking data and how that has impacted sports–perhaps that idea could be better unified and stated in a specific section of the article, or at the very least noted in the conclusion.

We hope that this revision addresses the reviewer's concerns about the pacing of the second half of the paper. In particular, we have added new subsections to both Sections 4 and 5, and added language tying together other parts of these Sections to ideas explored earlier in the paper.

Respectfully, we do not see player tracking data as a theme of the paper. However, one of the new subsections we added traces the evolution of data alongside the evolution in methodology and insights made in baseball. In the conclusion, we have pointed to this discussion as representative of similar evolutions happening in other sports. We hope that his will satisfy the reviewer's desire for clarity on this issue.

@@

> Pages 3,4, 5: The use of historical tables and data is powerful.  However, I worry about the quality of the images particularly Figure 2.

We share this concern. We are also uncertain about any licensing restrictions. We have left the images as is for now, and will wait direction from Wiley.

> Page 6, line 8:  "the effect may be overstated" is somewhat ambiguous.  Is it the

We have clarified our language on this subject. The reason for the comment was related to the increased resolution of player tracking data. We have made this reasoning explicit.

### 2022-11-01 sports_analytics.Rmd
close #39

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible.

@@

## Expected winning percentage

@@

## Case study in the evolution of tools and research: baseball

As the granularity of baseball data has evolved over time, so too have the statistical methodologies for modeling that data, and the tools for working with it. 
For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
Over time, the resolution of baseball data has improved to include play-by-play data, pitch-by-pitch data, and now player tracking data. 
The \pkg{Lahman} package [@R-Lahman] has provided seasonal data for years. 
The \pkg{retrosheet} package [@R-retrosheet] now provides access to the historical play-by-play data available from Retrosheet (this is a comprehensive version of what Lindsay collected for his research). 
The \pkg{pitchRx} package [@R-pitchRx] provides access to pitch-by-pitch data that fueled innovative research into catcher framing [@deshpande2017hierarchical], pitch values [@healey2019bayesian], and pitch classification [@sidle2018using].
The \pkg{baseballr} package [@R-baseballr] now provides access to player tracking data from Statcast.
These data have led to investigations into how defensive shifts affect batting performance [@bouzarth2021swing], as well as how launch angles affect the probability of hitting a home run [@marchi2018analyzing].

@@

This, these R packages enable research by making data more easily available. 
Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts.]

@@

# Conclusion

@@

In drawing together these three big ideas in sports analytics, we have also drawn attention to some computational tools for doing the work and opportunities to showcase that work. 
It is through these exchanges of ideas, tools, models, and data that analytics moves our collective understanding of sports forward.

>>>>

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible. 
Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper.

@@

## Expected winning percentage {#sec:wpct}

@@

## Case study in the evolution of tools and research: baseball {#sec:baseball}

As the granularity of baseball data has evolved over time, so too have the statistical methodologies for modeling that data, and the tools for working with it.

For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
These data, now available through the \pkg{Lahman} package [@R-Lahman], were sufficient to study broad trends in baseball, and led to insights such as the value of expected winning percentage (see Section \ref{sec:wpct}) and the importance of on-base percentage relative to batting average. 
These relatively simple insights fueled the "Moneyball" [@lewis2004moneyball] era revolution in sports analytics [@baumer2014sabermetric].

Over time, the resolution of baseball data has improved to include play-by-play data, pitch-by-pitch data, and now player tracking data.

The \pkg{retrosheet} package [@R-retrosheet] now provides access to the historical play-by-play data available from Retrosheet (this is a comprehensive version of what Lindsay collected for his research).
This play-by-play data allowed researchers to learn about strategies, like those that we discussed in Section \ref{sec:ev}.
In baseball, this deepened our understanding of bunting, stolen bases, handedness, batting order, and many other aspects of the game. 
Play-by-play data underlies much of the analysis in @Tango2007book.

The \pkg{pitchRx} package [@R-pitchRx] provides access to pitch-by-pitch data that fueled innovative research into catcher framing [@deshpande2017hierarchical], pitch values [@healey2019bayesian], and pitch classification [@sidle2018using].
Catcher framing is a notable example of a concept that scouts talked about for decades, but that could not be quantified by analysts until data of the appropriate resolution became available.

While play-by-play data allows us to make valuations *between plays*, player tracking data allows us to make valuations *within plays*. 
The \pkg{baseballr} package [@R-baseballr] now provides access to player tracking data from Statcast.
These data have led to investigations into how defensive shifts affect batting performance [@bouzarth2021swing], as well as how launch angles affect the probability of hitting a home run [@marchi2018analyzing].

@@

Thus, these R packages enable research by making data more easily available. 
Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts (see Section \ref{sec:opportunities}).]

@@

# Conclusion {#sec:conclusion}

@@

In drawing together these three big ideas in sports analytics, we have also drawn attention to some computational tools for doing the work and opportunities to showcase that work. 
Our discussion in Section \ref{sec:baseball} shows how the increased resolution of available data has catalyzed new research directions in baseball, but this same dynamic is playing out in all sports. 
It is through these exchanges of ideas, tools, models, and data that analytics moves our collective understanding of sports forward.
